---
title: "Practical Machine Learning - Course Project"
author: "Lucas Moraes"
output: html_document
---

# Load libraries

```{r results="hide", warning=FALSE, error=FALSE, message=FALSE}
library(caret)
library(randomForest)
library(xgboost)
library(dplyr)
library(VIM)
```


# Get the data

```{r}
data_dir = "./data"
training_url = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test_url = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
training_file = "pml-training.csv"
test_file = "pml-test.csv"
if (!file.exists(data_dir)) {
  dir.create(data_dir)
}
if (!file.exists(file.path(data_dir, training_file))) {
  download.file(training_url, destfile=file.path(data_dir, training_file))
}
if (!file.exists(file.path(data_dir, test_file))) {
  download.file(test_url, destfile=file.path(data_dir, test_file))
}
```

## Read the Data

Load the data into 2 different data frames

```{r}
train <- read.csv(file.path(data_dir, training_file), na.strings=c("NA","#DIV/0!", ""), stringsAsFactors = F)
test <- read.csv(file.path(data_dir, test_file), na.strings=c("NA","#DIV/0!", ""))
dim(train)
dim(test)
head(train)
```

The training data set contains 19622 observations and 160 variables, while the testing data set contains 20 observations and 160 variables. The "classe" variable in the training set is the outcome to predict. 

## Clean the data

Check if in the observations are present NA values or missing OBS that can raise errors/bias during the model training.

```{r}
sum(complete.cases(train))
```

Too few observations to have a correct training.

### Eliminate the columns with NA/missing values

Let's see colnames

```{r}
colnames(train)
plot(colMeans(is.na(train)))
```

There are columns with a lot of missing values.

We will retain only the columns without NA values

First covert all the data in NUMERIC form to coerce the empty factor to NA

```{r}
trainClasse = train$classe
trainRaw = train[, sapply(train, is.numeric)]
testRaw = test[, sapply(test, is.numeric)]
```

Remove columns with NA values

```{r}
trainFilter <- trainRaw[, colSums(is.na(trainRaw)) == 0]
# Attach Classe variable
trainFilter$classe = trainClasse
testFilter <- testRaw[, colSums(is.na(testRaw)) == 0]
```

Dimension

```{r}
dim(trainFilter)
dim(testFilter)
```

Removing other unuseful columns like username, timestamp and ID

```{r}
unwanted = !grepl("X|timestamp", colnames(trainFilter))
cols = colnames(trainFilter)[unwanted]
trainFilter = trainFilter %>%
  select(cols)
unwanted = !grepl("X|timestamp", colnames(testFilter))
cols = colnames(testFilter)[unwanted]
testFilter = testFilter %>%
  select(cols)
```

Get dimension of the filtered dataset

```{r}
dim(trainFilter)
dim(testFilter)
```

## Slice the data

We will slice the Training data into **Training** and **Validation** set using the 80-20 rule.

```{r}
set.seed(01312021) # Today's date
inTrain <- createDataPartition(trainFilter$classe, p=0.70, list=F)
trainData <- trainFilter[inTrain, ]
validationData <- trainFilter[-inTrain, ]
dim(trainData)
```

# Data modeling

We will fit a model using **Random Forest** and **XGBoost** (very popular in challange like kaggle.com) for several reasons:

1. With tree-based models, **you can safely ignore** predictors correlation issues

2. Zero- and Near Zero-Variance Predictors **does not** imply on tree-based models

3. As each feature is processed separately, and the possible splits of the data don’t depend on scaling, no preprocessing like normalization or standardization of features is needed for decision tree algorithms.

## Random forest

### Model

```{r}
#controlRf <- trainControl(method="cv", 5, allowParallel = TRUE)
modelRf <- train(classe ~ ., data=trainData, ntree = 100, method="rf", trControl = trainControl(method = "cv", number = 5,allowParallel = TRUE))
modelRf
```

### Performance of the model on the validation data set

```{r}
# Change the factors to avoid Error: data and reference should be factors with the same levels.
predict_rf <- predict(modelRf, validationData)
confMatRF <- confusionMatrix(as.factor(validationData$classe), predict_rf)
confMatRF$table
```

Calculate accuracy with a Random Forest Model
```{r}
rfAccuracy = confMatRF$overall[[1]]
rfAccuracy
```


Very accurate model to classify **classe** feature

## XGBoost

```{r}
controlXGB <- trainControl(method="cv", 5, allowParallel = TRUE)
modelXGB <- train(classe ~ ., data=trainData, method="xgbTree", trControl=controlXGB)
```

```{r}
modelXGB
```

### Performance of the model on the validation data set

```{r}
predict_XGB <- predict(modelXGB, validationData)
confMatXGB <- confusionMatrix(as.factor(validationData$classe), predict_XGB)
confMatXGB$table
```

Calculate accuracy with a XGBoost Model
```{r}
xgbAccuracy = confMatXGB$overall[[1]]
xgbAccuracy
```


With XGB we reach a better accuracy on validation data.

Only 2 mislabeled prediction A->B

# Compare models

```{r}
# collect resamples
model_results <- resamples(list(RF=modelRf, XGB=modelXGB))
# summarize the distributions
summary(model_results)
# boxplots of results
bwplot(model_results)
# dot plots of results
dotplot(model_results)
```

# Predict Test data with RF and XGB

```{r}
resultRf <- predict(modelRf, testFilter[, -length(names(testFilter))])
resultXGB <- predict(modelXGB, testFilter[, -length(names(testFilter))])
resultRf
resultXGB
confusionMatrix(resultRf, resultXGB)$table
```


Finally the model predict the TEST data in the same way, but we noticed that XGB works better with the training set
